---
title: "Presidential Approval"
author: "Ed Young"
date:  '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  html_document:
    css: styles.css
    highlight: null
    theme: null
  df_print: kable
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  cache = FALSE
)

library(here)
source(here("R", "load_packages.R"))

# --- Automatically wrap plots in styled div
knitr::knit_hooks$set(plot = function(x, options) {
  paste0('<div class="plot-frame">\n',
         knitr::hook_plot_md(x, options),
         '\n</div>')
})


```

```{r helpers, include=FALSE}

# Helper function to fit a LOESS curve with optional weights and 95% CIs

fit_loess <- function(data, span = 0.5, weight_col = NULL) {
  weights <- if (!is.null(weight_col)) data[[weight_col]] else NULL

  model <- loess(
    percentage ~ as.numeric(end_date),
    data = data,
    weights = weights,
    span = span
  )

  preds <- predict(
    model,
    newdata = data.frame(end_date = as.numeric(data$end_date)),
    se = TRUE
  )

  tibble(
    end_date = data$end_date,
    response = data$response,
    percentage = data$percentage,
    fit = preds$fit,
    se = preds$se.fit,
    ci_upper = preds$fit + 1.96 * preds$se.fit,
    ci_lower = preds$fit - 1.96 * preds$se.fit
  )
}

# Helper function to produce label points

get_latest_label_point <- function(loess_data) {
  loess_data %>%
    group_by(response) %>%
    arrange(desc(end_date), desc(fit)) %>%
    slice(1) %>%
    ungroup()
}

# Define pollsters used by the New York Times

selected_pollsters <- c(
  "Ipsos", "American Research Group", 
  "Beacon Research/Shaw & Company Research", "Gallup", "AtlasIntel",
  "Hart Research Associates/Public Opinion Strategies", "Quinnipiac University",
  "Emerson College", "CNN/SSRS", "Cygnal Political", "Public Opinion Strategies",
  "Marist College", "Marquette Law School", "Pew Research Center"
)

# --- Define consistent color palette for plots

# --- Fill colors (main identity)
approval_fills <- c("yes" = "#D73027", "no" = "#1C6EB9")

# --- Stroke (outline) colors
approval_strokes <- c("yes" = "#7F0000", "no" = "#08306B")



```

```{r download-data}

# Define the URL and the local file path

url <- "https://www.nytimes.com/newsgraphics/polls/approval/president.csv"
data_dir <- here("data")
file_path <- file.path(data_dir, "president.csv")

# Determine whether a download is needed:
# - Either the file doesn't exist...
# - Or the file exists but was last modified before today.

download_needed <- !file.exists(file_path) || 
  difftime(Sys.time(), file.info(file_path)$mtime, units = "hours") > 6

if (download_needed) {
  download.file(url, file_path, mode = "wb")
  message("Downloaded a fresh copy of president.csv.")
} else {
  message("The president.csv file is already up-to-date; no download needed.")
}
```

```{r read-data}

# --- Read the downloaded NYT presidential approval data

president <- readr::read_csv(file_path, show_col_types = FALSE)

# --- Convert end_date to Date format
president$end_date <- lubridate::mdy(president$end_date)

```

```{r date-diagnostic, eval=FALSE, echo=FALSE}

file.info(file_path)$mtime


# Check the structure of the dataset
glimpse(president)

# See the first and last dates in the data
summary(president$end_date)

# Optional: explicitly check latest date
latest <- max(president$end_date, na.rm = TRUE)
cat("Latest date in dataset:", latest, "\n")

# Visual histogram of poll counts by end date
ggplot(president, aes(x = end_date)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "white") +
  labs(title = "Distribution of Poll End Dates",
       x = "End Date",
       y = "Number of Polls") +
  theme_minimal()
```

# I am using the New York Times Presidential Approval Polls to look at trends over time.  

##  The first graph is the raw data with LOESS smoothing curves.

```{r raw}

# --- Reshape to long format for LOESS modeling
president_long <- pivot_longer(president, 
                               cols = c("yes", "no"), 
                               names_to = "response", 
                               values_to = "percentage")

# --- Fit LOESS models (fixed span, no weights)
loess_all <- bind_rows(
  fit_loess(filter(president_long, response == "yes"), span = 0.5),
  fit_loess(filter(president_long, response == "no"), span = 0.5)
)

# --- Extract label point for latest smoothed value
label_points <- get_latest_label_point(loess_all)

# --- Plot with LOESS smoothing and confidence ribbon
ggplot() +
  # Draw ribbon first, from loess_all
  geom_ribbon(
    data = loess_all,
    aes(x = end_date, ymin = ci_lower, ymax = ci_upper, fill = response, group = response),
    alpha = 0.3,
    color = NA
  ) +
  # Points and lines, also from loess_all
  geom_point(
    data = loess_all,
    aes(x = end_date, y = percentage, color = response),
    alpha = 0.6
  ) +
  geom_line(
    data = loess_all,
    aes(x = end_date, y = fit, color = response),
    linewidth = 1
  ) +
  geom_text(
    data = label_points,
    aes(x = end_date, y = fit, label = paste0("Latest: ", round(fit, 1), "%")),
    hjust = -0.1, vjust = -0.2, size = 3,
    show.legend = FALSE
  ) +
  scale_fill_manual(values = approval_fills) +
  scale_color_manual(values = approval_strokes) +
  labs(
    title = "All Polls with LOESS Smoothing (Fixed Span)",
    x = "Poll End Date",
    y = "Response (%)",
    color = "Response",
    fill = "Response"
  ) +
  theme_minimal() +
  xlim(min(loess_all$end_date), max(loess_all$end_date) + 15)

  
```

##  This second graph uses the same data, but, following the New York Times, I only include the following selected pollsters: Ipsos, American Research Group, Beacon Research/Shaw & Company Research, Gallup, AtlasIntel, Hart Research Associates/Public Opinion Strategies, Quinnipiac University, Emerson College, CNN/SSRS, Cygnal Political, Public Opinion Strategies, Marist College, Marquette Law School, Pew Research Center.  In addition, the data is weighted for recency, sample size and Probability Panel methodology. 

### Weights
\[
\text{Weight}_i =
0.3 \cdot e^{-\log(2) \cdot \frac{\text{days ago}_i}{14}} 
+ 0.3 \cdot \frac{\log(1 + \text{sample size}_i)}{\max_j \log(1 + \text{sample size}_j)} 
+ 0.2 \cdot \mathbb{1}_{\{\text{methodology}_i = \text{``Probability Panel''}\}} 
+ 0.2 \cdot \mathbb{1}_{\{\text{pollster}_i \in \text{Selected Pollsters}\}}
\]


```{r weighted}

# --- Set reference date for time weighting
latest_date <- max(president$end_date, na.rm = TRUE)

# --- Compute weights
president_weight <- president %>%
  mutate(
    # --- Time decay weight (half-life = 14 days)
    days_ago = as.numeric(latest_date - end_date),
    time_weight = exp(-log(2) * days_ago / 14),

    # --- Sample size weight: log-transformed and normalized
    size_weight = log1p(sample_size),
    size_weight = size_weight / max(size_weight, na.rm = TRUE),

    # --- Probability sample weight: 1 if method includes "Probability Panel" and not "Nonprobability"
    prob_weight = ifelse(
      grepl("Probability Panel", methodology, ignore.case = TRUE) &
      !grepl("Nonprobability Panel", methodology, ignore.case = TRUE),
      1, 0
    ),

    # --- Selected pollster weight: 1 if in trusted list, 0 otherwise
    selected_pollster_weight = ifelse(pollster %in% selected_pollsters, 1, 0),

    # --- Final composite weight
    weight = (0.2 * time_weight) +
             (0.2 * size_weight) +
             (0.3 * prob_weight) +
             (0.3 * selected_pollster_weight)
  )

# --- Reshape to long format for LOESS modeling
president_weight_long <- pivot_longer(president_weight, 
                                      cols = c("yes", "no"), 
                                      names_to = "response", 
                                      values_to = "percentage")

# --- Fit LOESS models for each response
loess_all <- bind_rows(
  fit_loess(filter(president_weight_long, response == "yes"), span = 0.5, weight_col = "weight"),
  fit_loess(filter(president_weight_long, response == "no"),  span = 0.5, weight_col = "weight")
)

# --- Extract label point for latest smoothed value
label_points <- get_latest_label_point(loess_all)

# --- Plot with LOESS smoothing and confidence ribbon
ggplot(loess_all, aes(x = end_date, y = percentage, color = response)) +
  geom_point(alpha = 0.3) +
  geom_line(aes(y = fit), linewidth = 1) +
  geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper, fill = response), alpha = 0.2, color = NA) +
  geom_text(data = label_points,
            aes(x = end_date, y = fit, label = paste0("Latest: ", round(fit, 1), "%")),
            hjust = -0.1, vjust = -0.2, size = 3, show.legend = FALSE) +
  scale_fill_manual(values = approval_fills) +
  scale_color_manual(values = approval_strokes) +
  labs(
    title = "All Pollsters Weighted by Recency, Sample Size, Method, and Credibility",
    x = "Poll End Date",
    y = "Response (%)",
    color = "Response",
    fill = "Response"
  ) +
  theme_minimal() +
  xlim(min(loess_all$end_date), max(loess_all$end_date) + 15)

```

# Here are some diagnostics to help think about these different factors

## First is a graph using only the Probability Panels

```{r Probability Diagnostic}
# --- Filter to only Probability Panel
president_prob <- president %>%
  filter(
    grepl("Probability Panel", methodology, ignore.case = TRUE) & 
    !grepl("Nonprobability Panel", methodology, ignore.case = TRUE)
  )

# --- Reshape to long format for LOESS modeling
president_long <- pivot_longer(president_prob, 
                               cols = c("yes", "no"), 
                               names_to = "response", 
                               values_to = "percentage")

# --- Fit LOESS models (fixed span, no weights)
loess_all <- bind_rows(
  fit_loess(filter(president_long, response == "yes"), span = 0.5),
  fit_loess(filter(president_long, response == "no"), span = 0.5)
)

# --- Extract label point for latest smoothed value
label_points <- get_latest_label_point(loess_all)

# --- Plot with LOESS smoothing and confidence ribbon
ggplot(loess_all, aes(x = end_date, y = percentage, color = response)) +
  geom_point(alpha = 0.6) +
  geom_line(aes(y = fit), linewidth = 1) +
  geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper, fill = response), alpha = 0.2, color = NA) +
  geom_text(data = label_points,
          aes(x = end_date, y = fit, label = paste0("Latest: ", round(fit, 1), "%")),
          hjust = -0.1, vjust = -0.2, size = 3, show.legend = FALSE) +
  scale_fill_manual(values = approval_fills) +
  scale_color_manual(values = approval_strokes) +
  labs(
    title = "Probability Panel Polls Only",
    x = "Poll End Date",
    y = "Response (%)",
    color = "Response",
    fill = "Response"
  ) +
  theme_minimal() +
  xlim(min(loess_all$end_date), max(loess_all$end_date) + 15)

```

## This is a graph using only selected pollsters

```{r selected}

# --- Filter to only Selected Pollsters
president_selected <- president %>%
  filter(pollster %in% selected_pollsters) 

# --- Reshape to long format for LOESS modeling
president_long <- pivot_longer(president_selected,
                               cols = c("yes", "no"), 
                               names_to = "response", 
                               values_to = "percentage")
# --- Fit LOESS models (fixed span, no weights)
loess_all <- bind_rows(
  fit_loess(filter(president_long, response == "yes"), span = 0.5),
  fit_loess(filter(president_long, response == "no"), span = 0.5)
)

# --- Extract label point for latest smoothed value
label_points <- get_latest_label_point(loess_all)


# --- Plot with LOESS smoothing and confidence ribbon
ggplot(loess_all, aes(x = end_date, y = percentage, color = response)) +
  geom_point(alpha = 0.6) +
  geom_line(aes(y = fit), linewidth = 1) +
  geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper, fill = response),
              alpha = 0.2, color = NA) +
  geom_text(data = label_points,
          aes(x = end_date, y = fit, label = paste0("Latest: ", round(fit, 1), "%")),
          hjust = -0.1, vjust = -0.2, size = 3, show.legend = FALSE) +
  scale_fill_manual(values = approval_fills) +
  scale_color_manual(values = approval_strokes) +
  labs(
    title = "Selected Pollsters Only",
    x = "Poll End Date",
    y = "Response (%)",
    color = "Response",
    fill = "Response"
  ) +
  theme_minimal() +
  xlim(min(loess_all$end_date), max(loess_all$end_date) + 15)

```

## This a graph weighted only by sample size

```{r Sample Size Diagnostic}

# --- Compute log-scaled, normalized sample size weight
president_size <- president %>%
  mutate(
    size_weight = log1p(sample_size),
    size_weight = size_weight / max(size_weight, na.rm = TRUE)
  )

# --- Reshape to long format for LOESS modeling
president_size_long <- pivot_longer(president_size, 
                                    cols = c("yes", "no"), 
                                    names_to = "response", 
                                    values_to = "percentage")

# --- Fit LOESS models (fixed span, weighted for size)
loess_all <- bind_rows(
  fit_loess(filter(president_size_long, response == "yes"), span = 0.5, weight_col = "size_weight"),
  fit_loess(filter(president_size_long, response == "no"),  span = 0.5, weight_col = "size_weight")
)

# --- Extract label point for latest smoothed value
label_points <- get_latest_label_point(loess_all)

# --- Plot with LOESS smoothing and confidence ribbon
ggplot() +
  geom_point(
    data = president_size_long,
    aes(x = end_date, y = percentage, color = response, fill = response, size = size_weight),
    shape = 21,
    stroke =0.4,
    alpha = 0.3
  ) +
  geom_line(data = loess_all,
            aes(x = end_date, y = fit, color = response), linewidth = 1) +
  geom_ribbon(data = loess_all,
              aes(x = end_date, ymin = ci_lower, ymax = ci_upper, fill = response), 
              alpha = 0.4, color = NA) +
  geom_text(data = label_points,
          aes(x = end_date, y = fit, label = paste0("Latest: ", round(fit, 1), "%")),
          hjust = -0.1, vjust = -0.2, size = 3, show.legend = FALSE) +
  scale_fill_manual(values = approval_fills) +
  scale_color_manual(values = approval_strokes) +
  scale_size_continuous(range = c(1, 5)) +
  labs(
    title = "Sample Size Only",
    x = "Poll End Date",
    y = "Response (%)",
    color = "Response",
    fill = "Response"
  ) +
  theme_minimal() +
  xlim(min(loess_all$end_date), max(loess_all$end_date) + 15)

```

## Polls weighted only for recency

```{r recency}
# --- Compute time decay weight (half-life = 14 days)
latest_date <- max(president$end_date, na.rm = TRUE)

president_recency <- president %>%
  mutate(
    days_ago = as.numeric(latest_date - end_date),
    time_weight = exp(-log(2) * days_ago / 14)
  )

# --- Reshape to long format for LOESS modeling
president_recency_long <- pivot_longer(president_recency,
                                       cols = c("yes", "no"),
                                       names_to = "response",
                                       values_to = "percentage")

# --- Fit LOESS models (fixed span, no weights)
loess_all <- bind_rows(
  fit_loess(filter(president_recency_long, response == "yes"), span = 0.5, weight_col = "time_weight"),
  fit_loess(filter(president_recency_long, response == "no"),  span = 0.5, weight_col = "time_weight")
)

# --- Extract label point for latest smoothed value
label_points <- get_latest_label_point(loess_all)


# --- Plot with LOESS smoothing and confidence ribbon
ggplot(loess_all, aes(x = end_date, y = percentage, color = response)) +
  geom_point(alpha = 0.3) +
  geom_line(aes(y = fit), size = 1) +
  geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper, fill = response), alpha = 0.2, color = NA) +
  geom_text(data = label_points,
          aes(x = end_date, y = fit, label = paste0("Latest: ", round(fit, 1), "%")),
          hjust = -0.1, vjust = -0.2, size = 3, show.legend = FALSE) +
  scale_fill_manual(values = approval_fills) +
  scale_color_manual(values = approval_strokes) +
  labs(
    title = "Recency Only",
    x = "Poll End Date",
    y = "Response (%)",
    color = "Response",
    fill = "Response"
  ) +
  theme_minimal() +
  xlim(min(loess_all$end_date), max(loess_all$end_date) + 15)

```

##  Plot weight vs. end_date to see how time decay is working:

```{r Time Decau Diagnostic}
ggplot(president_weight, aes(x = end_date, y = weight)) +
  geom_point(alpha = 0.7) +
  labs(title = "Poll Weights Over Time",
       x = "Poll End Date",
       y = "Computed Weight") +
  theme_minimal()
```

##  break down the contributions of time, sample size, and methodology to the final weight

```{r Variable Contribution Diagnostic}
president_weight_long <- president_weight %>%
  select(end_date, time_weight, size_weight, prob_weight, selected_pollster_weight) %>%
  pivot_longer(cols = -end_date, names_to = "component", values_to = "value")

ggplot(president_weight_long, aes(x = end_date, y = value, color = component)) +
  geom_line(stat = "summary", fun = mean) +
  labs(title = "Average Contribution of Weight Components Over Time",
       x = "Poll End Date", y = "Component Value",
       color = "Component") +
  theme_minimal()

# Visual histogram of poll counts by end date
ggplot(president_weight_long, aes(x = end_date)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "white") +
  labs(title = "Distribution of Poll End Dates",
       x = "End Date",
       y = "Number of Polls") +
  theme_minimal()
```