---
title: "Presidential Approval"
author: "Ed Young"
date:  '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  pdf_document: default
  df_print: kable
  html_document:
    css: styles.css
    highlight: null
    theme: null
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)

.old_plot_hook <- knitr::knit_hooks$get("plot")

knitr::knit_hooks$set(plot = function(x, options) {
  out <- .old_plot_hook(x, options)
  if (knitr::is_html_output()) {
    out <- paste0('<div class="plot-frame">\n', out, '\n</div>')
  }
  out
})

library(here)
source(here("R", "load_packages.R"))

```

```{r helpers, include=FALSE}

# Helper function to fit a LOESS curve with optional weights and 95% CIs.
# 
# Expected input columns: end_date (Date), percentage (numeric),
#   response (character), size_weight (numeric), sample_size (numeric).
#
# Arguments:
#   data       : data frame with the columns above
#   span       : numeric span (0-1), or NULL to auto-select via fANCOVA::loess.as()
#   weight_col : name of column to use as LOESS weights, or NULL for unweighted
#   criterion  : span selection criterion when span = NULL; "aicc" (default) or "gcv"

fit_loess <- function(data, span = NULL, weight_col = NULL, criterion = "aicc") {

  weights <- if (!is.null(weight_col)) data[[weight_col]] else NULL
  x <- as.numeric(data$end_date)
  y <- data$percentage

  if (is.null(span)) {
    # Auto-select span using fANCOVA; weights passed via ... to loess internally
    model <- if (!is.null(weights)) {
      fANCOVA::loess.as(x, y, criterion = criterion, weights = weights)
    } else {
      fANCOVA::loess.as(x, y, criterion = criterion)
    }
    selected_span <- model$pars$span
  } else {
    model <- loess(
      percentage ~ as.numeric(end_date),
      data     = data,
      weights  = weights,
      span     = span
    )
    selected_span <- span
  }

  preds <- predict(
    model,
    newdata = data.frame(end_date = x),
    se      = TRUE
  )

  tibble(
    end_date   = data$end_date,
    response   = data$response,
    percentage = data$percentage,
    size_weight = data$size_weight,
    sample_size = data$sample_size,
    fit        = preds$fit,
    se         = preds$se.fit,
    ci_upper   = preds$fit + 1.96 * preds$se.fit,
    ci_lower   = preds$fit - 1.96 * preds$se.fit,
    span_used  = selected_span
  )
}

# Helper function to produce label points

get_latest_label_point <- function(loess_data) {
  loess_data %>%
    group_by(response) %>%
    arrange(desc(end_date), desc(fit)) %>%
    slice(1) %>%
    ungroup()
}

# Define pollsters used by the New York Times

selected_pollsters <- c("Ipsos", 
                        "American Research Group",
                        "Beacon Research/Shaw & Company Research", 
                        "Gallup", 
                        "AtlasIntel",
                        "Hart Research Associates/Public Opinion Strategies", 
                        "Quinnipiac University",
                        "Emerson College", 
                        "CNN/SSRS", 
                        "Cygnal Political", 
                        "Public Opinion Strategies",
                        "Marist College", 
                        "Marquette Law School", 
                        "Pew Research Center",
                        "AP-NORC"
)

# --- Define consistent color palette for plots

# --- Fill colors (main identity)
approval_fills <- c("yes" = "#D73027", "no" = "#1C6EB9")

# --- Stroke (outline) colors
approval_strokes <- c("yes" = "#7F0000", "no" = "#08306B")

# Helper to set consistent x-axis padding so geom_text labels aren't clipped.
# Increase pad_days here to give labels more room.
approval_xlim <- function(data, pad_days = 40) {
  xlim(min(data$end_date), max(data$end_date) + pad_days)
}

```

```{r download-data}

# Define the URL and the local file path

url <- "https://www.nytimes.com/newsgraphics/polls/approval/president.csv"
data_dir <- here("data")
file_path <- file.path(data_dir, "president.csv")

# Determine whether a download is needed:
# - Either the file doesn't exist...
# - Or the file exists but was last modified before today.

download_needed <- !file.exists(file_path) || 
  difftime(Sys.time(), file.info(file_path)$mtime, units = "hours") > 6

if (download_needed) {
  download.file(url, file_path, mode = "wb")
  message("Downloaded a fresh copy of president.csv.")
} else {
  message("The president.csv file is already up-to-date; no download needed.")
}
```

```{r read-data}

# --- Read the downloaded NYT presidential approval data

president <- readr::read_csv(file_path, show_col_types = FALSE)

# --- Convert end_date to Date format
president$end_date <- lubridate::mdy(president$end_date)
president <- president %>% select(pollster, methodology, start_date, end_date, sample_size, population, yes, no, alternate_answers)

# Filter to probability samples that are *not* in the selected pollster list
prob_not_selected <- president %>%
  filter(
     (grepl("Probability Panel", methodology, ignore.case = TRUE) & !grepl("Nonprobability Panel", methodology, ignore.case = TRUE)),
    !pollster %in% selected_pollsters
  ) %>%
  distinct(pollster)

```

```{r date-diagnostic, eval=FALSE, echo=FALSE}

file.info(file_path)$mtime


# Check the structure of the dataset
glimpse(president)

# See the first and last dates in the data
summary(president$end_date)

# Optional: explicitly check latest date
latest <- max(president$end_date, na.rm = TRUE)
cat("Latest date in dataset:", latest, "\n")

# Visual histogram of poll counts by end date
ggplot(president, aes(x = end_date)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "white") +
  labs(title = "Distribution of Poll End Dates",
       x = "End Date",
       y = "Number of Polls") +
  theme_minimal()
```

# I am using the New York Times Presidential Approval Polls to look at trends over time.  

##  The first graph is the raw data with LOESS smoothing curves.

```{r raw}

# --- Reshape to long format for LOESS modeling
president_long <- pivot_longer(president, 
                               cols = c("yes", "no"), 
                               names_to = "response", 
                               values_to = "percentage") %>%
  mutate(
    size_weight = log1p(sample_size),
    size_weight = size_weight / max(size_weight, na.rm = TRUE)
  )

# --- Fit LOESS models (fixed span, no weights)
loess_all <- bind_rows(
  fit_loess(filter(president_long, response == "yes"), span = 0.5),
  fit_loess(filter(president_long, response == "no"), span = 0.5)
)

# --- Extract label point for latest smoothed value
label_points <- get_latest_label_point(loess_all)

# --- Plot with LOESS smoothing and confidence ribbon
ggplot() +
  # Draw ribbon first, from loess_all
  geom_ribbon(
    data = loess_all,
    aes(x = end_date, ymin = ci_lower, ymax = ci_upper, fill = response, group = response),
    alpha = 0.3,
    color = NA
  ) +
  # Points and lines, also from loess_all
  geom_point(
    data = loess_all,
    aes(x = end_date, y = percentage, color = response, size = sample_size),
    alpha = 0.1
  ) +
  geom_line(
    data = loess_all,
    aes(x = end_date, y = fit, color = response),
    linewidth = 1
  ) +
  geom_text(
    data = label_points,
    aes(x = end_date, y = fit, label = paste0(round(fit, 1), "%")),
    hjust = -0.1, vjust = -0.2, size = 3,
    show.legend = FALSE
  ) +
  scale_fill_manual(values = approval_fills) +
  scale_color_manual(values = approval_strokes) +
  labs(
    title = "All Polls with LOESS Smoothing (Fixed Span)",
    x = "Poll End Date",
    y = "Response (%)",
    color = "Response",
    fill = "Response",
    size = "Sample Size"
  ) +
  theme_minimal() +
  approval_xlim(loess_all)

  
```

##  This second graph uses the same data, but weighted for recency, sample size, Probability Panel methodology, polls of recent voters, and the use of selected pollsters. Following the New York Times, the selected pollsters are: Ipsos, American Research Group, Beacon Research/Shaw & Company Research, Gallup, AtlasIntel, Hart Research Associates/Public Opinion Strategies, Quinnipiac University, Emerson College, CNN/SSRS, Cygnal Political, Public Opinion Strategies, Marist College, Marquette Law School, Pew Research Center.  The weighting calculation is:

### Weights
\[
\begin{aligned}
\text{Weight}_i =\; & 0.2 \cdot e^{-\log(2) \cdot \frac{\text{days ago}_i}{14}} 
+ 0.2 \cdot \frac{\log(1 + \text{sample size}_i)}{\max_j \log(1 + \text{sample size}_j)} \\
& + 0.25 \cdot \mathbb{1}_{\{\text{methodology}_i = \text{``Probability Panel''}\}} 
+ 0.25 \cdot \mathbb{1}_{\{\text{pollster}_i \in \text{Selected Pollsters}\}}
+ 0.1 \cdot \mathbb{1}_{\{\text{methodology}_i = \text{``Recent Voters''}\}}
\end{aligned}
\]

```{r weighted}

# --- Set reference date for time weighting
latest_date <- max(president$end_date, na.rm = TRUE)

# --- Compute weights
president_weight <- president %>%
  mutate(
    # --- Time decay weight (half-life = 14 days)
    days_ago = as.numeric(latest_date - end_date),
    time_weight = exp(-log(2) * days_ago / 14),

    # --- Sample size weight: log-transformed and normalized
    size_weight = log1p(sample_size),
    size_weight = size_weight / max(size_weight, na.rm = TRUE),

    # --- Probability sample weight: 1 if method includes "Probability Panel" and not "Nonprobability"
    prob_weight = ifelse(
      grepl("Probability Panel", methodology, ignore.case = TRUE) &
      !grepl("Nonprobability Panel", methodology, ignore.case = TRUE),
      1, 0
    ),

    # --- Selected pollster weight: 1 if in trusted list, 0 otherwise
    selected_pollster_weight = ifelse(pollster %in% selected_pollsters, 1, 0),
    
    # --- Recent Voter weight:  1 if poll is of recent voters, 0 otherwise
    rv_weight = ifelse(population == "rv", 1,0),

    # --- Final composite weight
    weight = (0.2 * time_weight) +
             (0.2 * size_weight) +
             (0.25 * prob_weight) +
             (0.25 * selected_pollster_weight) +
             (0.1 * rv_weight)
  )

# --- Reshape to long format for LOESS modeling
president_weight_long <- pivot_longer(president_weight, 
                                      cols = c("yes", "no"), 
                                      names_to = "response", 
                                      values_to = "percentage")

# --- Fit LOESS models for each response
loess_all <- bind_rows(
  fit_loess(filter(president_weight_long, response == "yes"), span = 0.5, weight_col = "weight"),
  fit_loess(filter(president_weight_long, response == "no"),  span = 0.5, weight_col = "weight")
)

# --- Extract label point for latest smoothed value
label_points <- get_latest_label_point(loess_all)

# --- Plot with LOESS smoothing and confidence ribbon
ggplot() +
  geom_point(data = loess_all,
             aes(x = end_date, y = percentage, color = response, size = sample_size), 
             alpha = 0.1) +
  geom_line(data = loess_all,
            aes(x = end_date, y = fit, color = response), 
            linewidth = 1) +
  geom_ribbon(data = loess_all,
              aes(x = end_date, ymin = ci_lower, ymax = ci_upper, fill = response, group = response), 
              alpha = 0.2, 
              color = NA) +
  geom_text(data = label_points,
            aes(x = end_date, y = fit, label = paste0(round(fit, 1), "%")),
            hjust = -0.1, vjust = -0.2, size = 3, show.legend = FALSE) +
  scale_fill_manual(values = approval_fills) +
  scale_color_manual(values = approval_strokes) +
  guides(
  color = guide_legend(order = 1),
  fill  = guide_legend(order = 1),
  size  = guide_legend(order = 2)
  ) +
  labs(
    title = "All Pollsters Weighted by Recency, Sample Size, Method, Selected Pollsters, and Recent Voter Polls",
    x = "Poll End Date",
    y = "Response (%)",
    color = "Response",
    fill = "Response",
    size = "Sample Size"
  ) +
  theme_minimal() +
  approval_xlim(loess_all)

```

# Here are some diagnostics to help think about these different factors

## First is a graph using only the Probability Panels

```{r Probability Diagnostic}
# --- Filter to only Probability Panel
president_prob <- president %>%
  filter(
    grepl("Probability Panel", methodology, ignore.case = TRUE) & 
    !grepl("Nonprobability Panel", methodology, ignore.case = TRUE)
  ) %>%
  mutate(
    size_weight = log1p(sample_size),
    size_weight = size_weight / max(size_weight, na.rm = TRUE)
  )

# --- Reshape to long format for LOESS modeling
president_long <- pivot_longer(president_prob, 
                               cols = c("yes", "no"), 
                               names_to = "response", 
                               values_to = "percentage")

# --- Fit LOESS models (fixed span, no weights)
loess_all <- bind_rows(
  fit_loess(filter(president_long, response == "yes"), span = 0.5),
  fit_loess(filter(president_long, response == "no"), span = 0.5)
)

# --- Extract label point for latest smoothed value
label_points <- get_latest_label_point(loess_all)

# --- Plot with LOESS smoothing and confidence ribbon
ggplot() +
  geom_point(data = loess_all,
             aes(x = end_date, y = percentage, color = response, size = sample_size), alpha = 0.1) +
  geom_line(data = loess_all,
            aes(x = end_date, y = fit, color = response), linewidth = 1) +
  geom_ribbon(data = loess_all,
              aes(x = end_date, ymin = ci_lower, ymax = ci_upper, fill = response), alpha = 0.2, color = NA) +
  geom_text(data = label_points,
          aes(x = end_date, y = fit, label = paste0(round(fit, 1), "%")),
          hjust = -0.1, vjust = -0.2, size = 3, show.legend = FALSE) +
  scale_fill_manual(values = approval_fills) +
  scale_color_manual(values = approval_strokes) +
  guides(
  color = guide_legend(order = 1),
  fill  = guide_legend(order = 1),
  size  = guide_legend(order = 2)
  ) +
  labs(
    title = "Probability Panel Polls Only",
    x = "Poll End Date",
    y = "Response (%)",
    color = "Response",
    fill = "Response",
    size = "Sample Size"
  ) +
  theme_minimal() +
  approval_xlim(loess_all)

```

## This is a graph using only polls of likely voters

```{r Likely Voters}
# --- Filter to only Probability Panel
president_lv <- president %>%
  filter(population == "lv"  ) %>%
  mutate(
    size_weight = log1p(sample_size),
    size_weight = size_weight / max(size_weight, na.rm = TRUE)
  )

# --- Reshape to long format for LOESS modeling
president_long <- pivot_longer(president_lv, 
                               cols = c("yes", "no"), 
                               names_to = "response", 
                               values_to = "percentage")

# --- Fit LOESS models (fixed span, no weights)
loess_all <- bind_rows(
  fit_loess(filter(president_long, response == "yes"), span = 0.5),
  fit_loess(filter(president_long, response == "no"), span = 0.5)
)

# --- Extract label point for latest smoothed value
label_points <- get_latest_label_point(loess_all)

# --- Plot with LOESS smoothing and confidence ribbon
ggplot() +
  geom_point(data = loess_all,
             aes(x = end_date, y = percentage, color = response, size = sample_size),
             alpha = 0.1) +
  geom_line(data = loess_all,
            aes(x = end_date, y = fit, color = response), 
            linewidth = 1) +
  geom_ribbon(data = loess_all,
              aes(x = end_date, ymin = ci_lower, ymax = ci_upper, fill = response), 
              alpha = 0.2, 
              color = NA) +
  geom_text(data = label_points,
          aes(x = end_date, y = fit, label = paste0(round(fit, 1), "%")),
          hjust = -0.1, vjust = -0.2, size = 3, show.legend = FALSE) +
  scale_fill_manual(values = approval_fills) +
  scale_color_manual(values = approval_strokes) +
  guides(
  color = guide_legend(order = 1),
  fill  = guide_legend(order = 1),
  size  = guide_legend(order = 2)
  ) +
  labs(
    title = "Polls of Likely Voters Only",
    x = "Poll End Date",
    y = "Response (%)",
    color = "Response",
    fill = "Response",
    size = "Sample Size"
  ) +
  theme_minimal() +
  approval_xlim(loess_all)

```

## This is a graph of polls of recent voters only

```{r recent Voters}
# --- Filter to only Probability Panel
president_lv <- president %>%
  filter(population == "rv"  ) %>%
  mutate(
    size_weight = log1p(sample_size),
    size_weight = size_weight / max(size_weight, na.rm = TRUE)
  )

# --- Reshape to long format for LOESS modeling
president_long <- pivot_longer(president_lv, 
                               cols = c("yes", "no"), 
                               names_to = "response", 
                               values_to = "percentage")

# --- Fit LOESS models (fixed span, no weights)
loess_all <- bind_rows(
  fit_loess(filter(president_long, response == "yes"), span = 0.5),
  fit_loess(filter(president_long, response == "no"), span = 0.5)
)

# --- Extract label point for latest smoothed value
label_points <- get_latest_label_point(loess_all)

# --- Plot with LOESS smoothing and confidence ribbon
ggplot() +
  geom_point(data = loess_all,
             aes(x = end_date, y = percentage, color = response, size = sample_size), alpha = 0.1) +
  geom_line(data = loess_all,
            aes(x = end_date, y = fit, color = response), linewidth = 1) +
  geom_ribbon(data = loess_all,
              aes(x = end_date, ymin = ci_lower, ymax = ci_upper, fill = response), alpha = 0.2, color = NA) +
  geom_text(data = label_points,
          aes(x = end_date, y = fit, label = paste0(round(fit, 1), "%")),
          hjust = -0.1, vjust = -0.2, size = 3, show.legend = FALSE) +
  scale_fill_manual(values = approval_fills) +
  scale_color_manual(values = approval_strokes) +
  guides(
  color = guide_legend(order = 1),
  fill  = guide_legend(order = 1),
  size  = guide_legend(order = 2)
  ) +
  labs(
    title = "Polls of Recent Voters Only",
    x = "Poll End Date",
    y = "Response (%)",
    color = "Response",
    fill = "Response",
    size = "Sample Size"
  ) +
  theme_minimal() +
  approval_xlim(loess_all)

```

## This is a graph using only selected pollsters

```{r selected}

# --- Filter to only Selected Pollsters
president_selected <- president %>%
  filter(pollster %in% selected_pollsters) %>%
  mutate(
    size_weight = log1p(sample_size),
    size_weight = size_weight / max(size_weight, na.rm = TRUE)
  )

# --- Reshape to long format for LOESS modeling
president_long <- pivot_longer(president_selected,
                               cols = c("yes", "no"), 
                               names_to = "response", 
                               values_to = "percentage")
# --- Fit LOESS models (fixed span, no weights)
loess_all <- bind_rows(
  fit_loess(filter(president_long, response == "yes"), span = 0.5),
  fit_loess(filter(president_long, response == "no"), span = 0.5)
)

# --- Extract label point for latest smoothed value
label_points <- get_latest_label_point(loess_all)


# --- Plot with LOESS smoothing and confidence ribbon
ggplot() +
  geom_point(data = loess_all,
             aes(x = end_date, y = percentage, color = response, size = sample_size), alpha = 0.1) +
  geom_line(data = loess_all,
            aes(x = end_date, y = fit, color = response), linewidth = 1) +
  geom_ribbon(data = loess_all,
              aes(x = end_date, ymin = ci_lower, ymax = ci_upper, fill = response),
              alpha = 0.2, color = NA) +
  geom_text(data = label_points,
          aes(x = end_date, y = fit, label = paste0(round(fit, 1), "%")),
          hjust = -0.1, vjust = -0.2, size = 3, show.legend = FALSE) +
  scale_fill_manual(values = approval_fills) +
  scale_color_manual(values = approval_strokes) +
  guides(
  color = guide_legend(order = 1),
  fill  = guide_legend(order = 1),
  size  = guide_legend(order = 2)
  ) +
  labs(
    title = "Selected Pollsters Only",
    x = "Poll End Date",
    y = "Response (%)",
    color = "Response",
    fill = "Response",
    size = "Sample Size"
  ) +
  theme_minimal() +
  approval_xlim(loess_all)

```

## This a graph weighted only by sample size

```{r Sample Size Diagnostic}

# --- Compute log-scaled, normalized sample size weight
president_size <- president %>%
  mutate(
    size_weight = log1p(sample_size),
    size_weight = size_weight / max(size_weight, na.rm = TRUE)
  )

# --- Reshape to long format for LOESS modeling
president_size_long <- pivot_longer(president_size, 
                                    cols = c("yes", "no"), 
                                    names_to = "response", 
                                    values_to = "percentage")

# --- Fit LOESS models (fixed span, weighted for size)
loess_all <- bind_rows(
  fit_loess(filter(president_size_long, response == "yes"), span = 0.5, weight_col = "size_weight"),
  fit_loess(filter(president_size_long, response == "no"),  span = 0.5, weight_col = "size_weight")
)

# --- Extract label point for latest smoothed value
label_points <- get_latest_label_point(loess_all)

# --- Plot with LOESS smoothing and confidence ribbon
ggplot() +
  geom_point(
    data = president_size_long,
    aes(x = end_date, y = percentage, color = response, fill = response, size = sample_size),
    shape = 21,
    stroke =0.4,
    alpha = 0.1
  ) +
  geom_line(data = loess_all,
            aes(x = end_date, y = fit, color = response), linewidth = 1) +
  geom_ribbon(data = loess_all,
              aes(x = end_date, ymin = ci_lower, ymax = ci_upper, fill = response), 
              alpha = 0.4, color = NA) +
  geom_text(data = label_points,
          aes(x = end_date, y = fit, label = paste0(round(fit, 1), "%")),
          hjust = -0.1, vjust = -0.2, size = 3, show.legend = FALSE) +
  scale_fill_manual(values = approval_fills) +
  scale_color_manual(values = approval_strokes) +
  scale_size_continuous(range = c(1, 10)) +
  guides(
  color = guide_legend(order = 1),
  fill  = guide_legend(order = 1),
  size  = guide_legend(order = 2)
  ) +
  labs(
    title = "Sample Size Only",
    x = "Poll End Date",
    y = "Response (%)",
    color = "Response",
    fill = "Response",
    size = "Sample Size"
  ) +
  theme_minimal() +
  approval_xlim(loess_all)

```

## Polls weighted only for recency

```{r recency}
# --- Compute time decay weight (half-life = 14 days)
latest_date <- max(president$end_date, na.rm = TRUE)

president_recency <- president %>%
  mutate(
    days_ago = as.numeric(latest_date - end_date),
    time_weight = exp(-log(2) * days_ago / 14),
    size_weight = log1p(sample_size),
    size_weight = size_weight / max(size_weight, na.rm = TRUE)
  )

# --- Reshape to long format for LOESS modeling
president_recency_long <- pivot_longer(president_recency,
                                       cols = c("yes", "no"),
                                       names_to = "response",
                                       values_to = "percentage")

# --- Fit LOESS models (fixed span, no weights)
loess_all <- bind_rows(
  fit_loess(filter(president_recency_long, response == "yes"), span = 0.5, weight_col = "time_weight"),
  fit_loess(filter(president_recency_long, response == "no"),  span = 0.5, weight_col = "time_weight")
)

# --- Extract label point for latest smoothed value
label_points <- get_latest_label_point(loess_all)


# --- Plot with LOESS smoothing and confidence ribbon
ggplot(loess_all, aes(x = end_date, y = percentage, color = response)) +
  geom_point(alpha = 0.3) +
  geom_line(aes(y = fit), size = 1) +
  geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper, fill = response), alpha = 0.2, color = NA) +
  geom_text(data = label_points,
          aes(x = end_date, y = fit, label = paste0(round(fit, 1), "%")),
          hjust = -0.1, vjust = -0.2, size = 3, show.legend = FALSE) +
  scale_fill_manual(values = approval_fills) +
  scale_color_manual(values = approval_strokes) +
  guides(
  color = guide_legend(order = 1),
  fill  = guide_legend(order = 1),
  size  = guide_legend(order = 2)
  ) +
  labs(
    title = "Recency Only",
    x = "Poll End Date",
    y = "Response (%)",
    color = "Response",
    fill = "Response",
    size = "Sample Size"
  ) +
  theme_minimal() +
  approval_xlim(loess_all)

```

##  break down the contributions of time, sample size, and methodology to the final weight

```{r Variable Contribution Diagnostic}
president_weight_long <- president_weight %>%
  select(end_date, time_weight, size_weight, prob_weight, selected_pollster_weight) %>%
  pivot_longer(cols = -end_date, names_to = "component", values_to = "value")

ggplot(president_weight_long, aes(x = end_date, y = value, color = component)) +
  geom_line(stat = "summary", fun = mean) +
  labs(title = "Average Contribution of Weight Components Over Time",
       x = "Poll End Date", y = "Component Value",
       color = "Component") +
  theme_minimal()

# Visual histogram of poll counts by end date
ggplot(president_weight_long, aes(x = end_date)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "white") +
  labs(title = "Distribution of Poll End Dates",
       x = "End Date",
       y = "Number of Polls") +
  theme_minimal()
```